---
title: "2019 CrossFit Open, The Numbers Behind Participation"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=F, warning=F, include=F}
library(tidyverse)
library(lemon)
knit_print.data.frame <- lemon_print
```


```{r, message=F, warning=F, include=F}
men <- read_csv('/Users/silv6928/Data/athletes_men_2019_2019-02-27.csv')
women <- read_csv('/Users/silv6928/Data/athletes_women_2019_2019-02-27.csv')
# women csv had NULL gender so adjusted
women$gender <- 'F'

# Scores are empty because the Open has started
# Load both csvs for scores
scores_men <- read_csv('/Users/silv6928/Data/scores_men_2019_2019-02-27.csv')
scores_women <- read_csv('/Users/silv6928/Data/scores_women_2019_2019-02-27.csv')

athletes_2019 <- union(men, women)
scores_2019 <- union(scores_men, scores_women)
rm(men,women, scores_men, scores_women)

athletes_2019$affiliateName[athletes_2019$affiliateId==0] <- 'Unaffiliated'
```

```{r, message=F, warning=F, include=F}
athletes_2018 <- read_csv("/Users/silv6928/Data/athletes_2018.csv")
scores_2018 <- read_csv("/Users/silv6928/Data/scores_2018.csv")
```

```{r}
men <- read_csv("/Users/silv6928/Data/athletes_men_2017_2019-02-28.csv")
women <- read_csv('/Users/silv6928/Data/athletes_women_2017_2019-02-28.csv')

athletes_2017 <- union(men, women)
rm(men,women)

colnames(athletes_2017)[2] <- 'competitorId'
```

Below is an example the pre-processed data frames of the 2019 athlete data 
```{r, render=lemon_print}
head(athletes_2019[1:4,c(1:4,6)])
```

I used that data to build out the analysis in this blog on finding the true numbers behind the CrossFit Open participation. 

#### How many people have signed up in 2018 vs. 2019?

```{r}
num_athletes_18 <- length(unique(athletes_2018$competitorId))
num_athletes_18
```
We can also do the same thing for the 2019 data. And below you can see the number of sign ups in 2019.
```{r}
num_athletes_19 <- length(unique(athletes_2019$competitorId))
num_athletes_19
```
Next, we can actually can see the percentage change in total Open participation from 2018 to 2019. 
```{r}
(1-round(num_athletes_19/num_athletes_18,4))*100
```

```{r}
# Join 2018 and 2019 together by competitorId
athletes <- athletes_2019 %>%
  left_join(athletes_2018, by='competitorId')  %>%
  mutate(IsNew=ifelse(is.na(competitorName.y),1,0)) # mark if an athlete is New
```

Below is the number of New Athletes that signed up.
```{r}
# New Athletes
sum(athletes$IsNew)
```
Below is the number of Repeat Athletes that signed up.
```{r}
# Repeat Atheltes
num_athletes_19 - sum(athletes$IsNew)
```
We can then take the repeat athletes and determine, of those what's the percentage of people who signed up last year that also signed up this year?
```{r, warning=F}
round((num_athletes_19 - sum(athletes$IsNew))/num_athletes_18,4)*100
```

```{r}
sum(athletes$IsNew)/num_athletes_19
```


```{r}
athletes_2019 %>% 
  group_by(competitorId) %>%
  summarize(counts = n()) %>%
  filter(counts > 1)
```


```{r}
num_athletes_17 <- length(unique(athletes_2017$competitorId))
athletes_join <- athletes_2018 %>%
  left_join(athletes_2017, by='competitorId')  %>%
  mutate(IsNew=ifelse(is.na(name),1,0)) # mark if an athlete is New
```

```{r}
?
round((num_athletes_18 - sum(athletes_join$IsNew))/num_athletes_17,4)*100
```

```{r}
x = c(num_athletes_18 - sum(athletes_join$IsNew), num_athletes_19 - sum(athletes$IsNew))
n = c(num_athletes_18, num_athletes_19)
prop.test(x,n)
```

